---
title: "Midterm One"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
```

In this midterm we will analyze some data on the conservation status of species in North America and spending under the Endangered Species Act. 

Answer the following questions by using chunks of R code. Comment on what your code does. Make sure to add informative axis titles and, where appropriate, units to your answers. Upload the R markdown file and knitted output to Canvas. 

We will use the file `conservationdata.csv`. This dataset has information on North American species. It has five variables that are described in the table below.


```{r, echo=FALSE}
tab <- cbind(Name=c("speciesid", "speciesname", "taxon", "conservation status", "listed"), Description = c("unique ID", "scientific name", "Species group", "Conservation status in North America, according to   [NatureServe](https://www.natureserve.org/conservation-tools/conservation-rank-calculator): 1 = Critically Imperiled; 2 = Imperiled; 3 = Vulnerable; 4 = Apparently Secure; 5 = Secure; UNK = Unknown; Prob. Extinct = Probably Extinct; Extinct", "Is the species listed as threatened or endangered under the US Endangered Species Act: 0 = No; 1 = Yes"))

knitr::kable(tab, caption ='Table 1. Variables in "consevationdata.csv"' )
```


Read the file `conservationdata.csv`

```{r}
conservation_data <- read.csv("conservationdata.csv")
```

1. What fraction of species in the dataset are listed under the Endangered Species Act? (2 points)
> fraction_endangered <- mean(conservation_data$listed)
> fraction_endangered
[1] 0.0301353

2. Show how many (absolute and relative) species there are for each taxonomic group by making a data.frame in which the first column has the taxonomic groups, the second column is the number of species in that group, and the third column is the percentage of all species in that group.
> taxonomic_counts <- table(conservation_data$taxon)
> taxonomic_df <- data.frame(
+     Taxonomic_Group = names(taxonomic_counts),
+     Number_of_Species = as.numeric(taxonomic_counts),
+     Percentage_of_Total = (as.numeric(taxonomic_counts) / nrow(conservation_data)) * 100
+ )
> taxonomic_df
  Taxonomic_Group Number_of_Species Percentage_of_Total
1      Amphibians               319           0.5945059
2           Birds               795           1.4816057
3          Fishes              1453           2.7078907
4           Fungi              6270          11.6851169
5   Invertebrates             24407          45.4862276
6         Mammals               474           0.8833725
7          Plants             19511          36.3617727
8        Protists                79           0.1472287
9        Reptiles               350           0.6522793

3a) One interesting question is how the conservation status varies between different taxonomic groups. Make a plot showing the relative distribution of conservation status within each taxonomic group. There should be descriptive legend (with words, not with the numeric codes) (3 points)

You can use a "base" plotting method, or ggplot. 

If you are using ggplot, stat="count" (counts up and plots the number of observations, i.e. species, within each group) and position="fill" might both be useful.

> library(ggplot2)
> ggplot(conservation_data, aes(x = taxon, fill = factor(`conservation_status`))) +
+     geom_bar(position = "fill") +
+     labs(title = "Relative Distribution of Conservation Status by Taxonomic Group",
+          x = "Taxonomic Group",
+          y = "Proportion",
+          fill = "Conservation Status") +
+     theme_minimal()

3b) Based on this graph, what is something we might be concerned about in terms of analyzing the data on conservation status, particularly for fungi and invertebrates? (1 point)

**Answer:Based on the graph, one concern might be that there are very high proportion of UNK observations for fungi and invertebrates, making it difficult to draw robust conclusions about the distribution of conservation status for these groups.** 

Read in the second data file: `spendingdata.csv`

This dataset has a species ID that matches the species ID in the conservation dataset (speciesid), year, and the spending on conservation of that species (expressed in in 2015 dollars, i.e., accounting for inflation)

>spending_data <- read.csv("spendingdata.csv")

4a) Make a plot showing the distribution of spending in the year 2016 (3 points)

> ggplot(spending_data, aes(x = spending)) +
+     geom_histogram(binwidth = 1000000, fill = "blue", color = "black") +
+     labs(title = "Distribution of Spending in 2016",
+          x = "Spending (in millions)",
+          y = "Frequency") +
+     theme_minimal()

4b) Notice the (very) long right tail on spending data - we spend a lot on a very small number of species. Show the IDs of the 3 species with the most spending in 2016. (2 points)

> top_3_spending <- head(spending_data[order(spending_data$spending, decreasing = TRUE), ], 3)
> top_3_spending
     speciesid Year  spending
2111      1632 2011 313188886
2109      1632 2009 303991515
4642      4486 2012 294530155

5. Merge in the data from the conservation status data frame to the spending data frame, so that we have information on species names, taxonomic group, and conservation status with the spending data. (2 points); and use that to show the scientific names of the three species identified above.

> merged_data <- merge(spending_data, conservation_data, by = "speciesid")
> merged_data[merged_data$speciesid %in% top_3_spending$speciesid, c("speciesid", "speciesname")]
     speciesid              speciesname
2189      1632 Oncorhynchus tshawytscha
2190      1632 Oncorhynchus tshawytscha
2191      1632 Oncorhynchus tshawytscha
2192      1632 Oncorhynchus tshawytscha
2193      1632 Oncorhynchus tshawytscha
2194      1632 Oncorhynchus tshawytscha
2195      1632 Oncorhynchus tshawytscha
2196      1632 Oncorhynchus tshawytscha
2197      1632 Oncorhynchus tshawytscha
2198      1632 Oncorhynchus tshawytscha
2199      1632 Oncorhynchus tshawytscha
2200      1632 Oncorhynchus tshawytscha
2201      1632 Oncorhynchus tshawytscha
2202      1632 Oncorhynchus tshawytscha
2203      1632 Oncorhynchus tshawytscha
2204      1632 Oncorhynchus tshawytscha
2205      1632 Oncorhynchus tshawytscha
2206      1632 Oncorhynchus tshawytscha
2207      1632 Oncorhynchus tshawytscha
2208      1632 Oncorhynchus tshawytscha
2209      1632 Oncorhynchus tshawytscha
4742      4486      Oncorhynchus mykiss
4743      4486      Oncorhynchus mykiss
4744      4486      Oncorhynchus mykiss
4745      4486      Oncorhynchus mykiss
4746      4486      Oncorhynchus mykiss
4747      4486      Oncorhynchus mykiss
4748      4486      Oncorhynchus mykiss
4749      4486      Oncorhynchus mykiss
4750      4486      Oncorhynchus mykiss
4751      4486      Oncorhynchus mykiss
4752      4486      Oncorhynchus mykiss
4753      4486      Oncorhynchus mykiss
4754      4486      Oncorhynchus mykiss
4755      4486      Oncorhynchus mykiss
4756      4486      Oncorhynchus mykiss
4757      4486      Oncorhynchus mykiss
4758      4486      Oncorhynchus mykiss
4759      4486      Oncorhynchus mykiss
4760      4486      Oncorhynchus mykiss

Look up these scientific names - what is the common name for these species?

**Answer:Oncorhynchus tshawytscha, Oncorhynchus mykiss** 

6. Finally, we will use a regression to look at the relationship between spending and species taxon.

Because the distribution of spending is very right-skewed, it would be a good idea to take the logarithm of spending before using it in a regression. 

Remember that log(0)=infinity. That means we have to drop observations with zero spending before taking the logarithm.

a) Drop the rows where spending == 0 from the data frame and then make a new column with the logarithm (log()) of spending in each year. (2 points)

> log_spending_data <- subset(spending_data, spending > 0)
> log_spending_data$log_spending <- log(log_spending_data$spending)

Optional: Look at the distribution of the logged spending variable and see how it looks different from the plot you made in question 4a

b) Run a regression of logged spending on taxonomic group and print the summary for the regression below (3 points)

> merged_data$log_spending <- log(merged_data$spending + 1e-10)
> model <- lm(log_spending ~ taxon, data = merged_data)
> summary(model)

Call:
lm(formula = log_spending ~ taxon, data = merged_data)

Residuals:
    Min      1Q  Median      3Q     Max 
-35.145  -0.524   0.795   2.138   8.338 

Coefficients:
                   Estimate Std. Error t value Pr(>|t|)    
(Intercept)         11.1800     0.2569  43.514  < 2e-16 ***
taxonBirds           0.7404     0.2857   2.591  0.00956 ** 
taxonFishes          0.5221     0.2780   1.878  0.06043 .  
taxonFungi          -1.1748     0.8794  -1.336  0.18158    
taxonInvertebrates  -0.7499     0.2688  -2.790  0.00527 ** 
taxonMammals         0.9389     0.2894   3.244  0.00118 ** 
taxonPlants         -2.4709     0.2606  -9.480  < 2e-16 ***
taxonReptiles        0.3592     0.3273   1.097  0.27244    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 5.45 on 27622 degrees of freedom
Multiple R-squared:  0.05782,	Adjusted R-squared:  0.05758 
F-statistic: 242.1 on 7 and 27622 DF,  p-value: < 2.2e-16

c) The way to interpret these coefficients are as the fractional difference in spending between the taxonomic group (e.g. Birds, Fishes etc) and the "dropped" group, where by default the dropped group will be Amphibians. Positive numbers indicate that group has more spent on it than Amphibians and negative numbers indicate it has less spent on it. 

Based on your results in b, do we see statistically significant differences in spending between different taxonomic groups? If so, which kinds of species tend to have more spent on them and which have less? (1 points)

**Answer:Birds, Invertebrates, and Mammals tend to have significantly more spending. Fungi, Reptiles, and Plants do not show significant differences in spending have less spending compared to the reference group.** 

7. Push your R markdown file to your Github repository (2 points)
